BOLD Parameter Estimation Using Particle Filters

### 2010-08-22 doc/thesis/background.tex
small fixes, typos, and a strangely copied line

\begin{figure}
\includegraphics[width=16cm]{images/kalman}
%\includegraphics[trim=6cm .75cm 6cm .75cm,width=16cm]{images/gauss_step_point1sec_3sigma.pdf}
\caption{Example updates of a distribution using the Kalman Filter, \cite{Thrun2005}}
\label{fig:EKFWorking}
\end{figure}

@@

In Vakorin et al., a combination of Genetic Algorithms and Simulated Annealing were
used to estimate not only the parameters, but the true stimuli driving the BOLD
signal. This addresses the inherent uncertainty of exactly where and when 
stimuli actually get applied. Unfortunately this algorithm was extremely slow.

>>>>

\begin{figure}
\includegraphics[width=16cm]{images/kalman}
%\includegraphics[trim=6cm .75cm 6cm .75cm,width=16cm]{images/gauss_step_point1sec_3sigma.pdf}
\caption{Example updates of a distribution using the Kalman Filter.}
\label{fig:EKFWorking}
\end{figure}

@@

In Vakorin et al., a combination of Genetic Algorithms and Simulated Annealing were
used to estimate not only the parameters, but the true stimuli driving the BOLD
signal \cite{Vakorin2007}. 
This addresses the inherent uncertainty of exactly where and when 
stimuli actually get applied. Unfortunately this algorithm was extremely slow.

### 2010-08-22 doc/thesis/particlefilters.tex
FINAL!
small fixes, typos, and a strangely copied line

I will also make a few  simplifying assumptions for this work. 
First, the systems are assumed to be 
time invariant. This 
assumption is based on the idea that if you paused the system for $\Delta t$
seconds, when unfrozen the system would continue as if nothing happened. 
Few biological systems are predictable enough for them to be summarized
by a time varying function, least of all the brain. While heart beats are certainly
periodic and have an effect on the BOLD signal, the period varies too much
for the system to be considered varying with time. 
Next, its assumes that input cannot directly
influence the output, which in the case of the BOLD signal is a good assumption.
Finally, because the only difference between the members of $X(t)$ and 
$\Theta$ is an update function, from now on $X$ will encapsulate
$\Theta$. The assumptions now allow for a simplified version of the
state space equations:

\begin{eqnarray}
\dot{X}_k = f(X_{k-1}, u_k, \nu_x)
\label{eq:stateass}\\
Y_k = g(X_k, \nu_y)
\label{eq:measass}
\end{eqnarray}

@@

\begin{algorithm}
\caption{Resampling Algorithm}
\begin{algorithmic}
\STATE Calculate total weight, $W = \sum_{i=0}^{N_p} w^i$
\FORALL{$0 < i < N_p$}
    \STATE Draw $V$ from uniform range $[0, W]$
    \STATE $C = W_t$
    \FORALL{$0 < j < N_p$ and $C < V$}
        \STATE $C = C - w^j$
    \ENDFOR
    \STATE Add $[x^j, \frac{1}{N_p}]$ to the new distribution
\ENDFOR
\STATE 
\end{algorithmic}
\label{alg:Resampling}
\end{algorithm}
For infinite particles this new distribution will match the old.
Unfortunately, this isn't the truth in practice: since the support is
still limited to the original particles, the number of \emph{unique} particles can only go down.
This effect, dubbed particle impoverishment can result in excessive quantization
errors in the final distribution. However, there is a solution. Instead of sampling from the
discrete distribution, a smoothing kernel is applied, and particles are drawn from
that distribution. Because it is continuous, particle impoverishment
cannot occur. The easiest way to sample from the continuous distribution is to break the 
re-sampling down into two steps. After calculating an estimate of the scale of the original
distribution, algorithm \autoref{alg:Resampling} is performed. Next, a distribution is generated
based on the variance of the original distributions.
Finally, for each particle in the discretely re-sampled distribution, a sample is drawn from 
the smoothing 
distribution and added to the particle.  The regularization process is defined as:

@@

\STATE At $t + \Delta t$, $t \in T$, $P(x(t+\Delta t)) \approx 
\sum_{i=1}^{N_p} w_i(t)\delta\left(x - (x_i(t) + \int_t^{t+\Delta t} f(x(\tau), u(\tau)) d\tau) \right)$
 \end{algorithmic}
 \end{algorithm}

@@

With regularized resampling, it is possible to prevent both
particle degeneracy as well as particle impoverishment. Additionally,
as the particle filter converges, the density of particles
in the area of the solution goes up. This has a similar effect to 
simulated annealing where, as the algorithm approaches the end, the
random steps get smaller and smaller. Note however, that resampling
carries certain risks.  If for some reason the solution is not covered by the 
new support, the algorithm may not be able to reach the true value. 
The ultimate effect of this regularized resampling is a convergence similar to simulated annealing
or a genetic algorithm. Versions of $x$ that are fit (give good measurements) spawn more children 
nearby which allow for more accurate estimation near points of high likelihood. 
As the variance of the estimated
$x$'s decrease, the radius in which children are spawned also decreases. Eventually the radius
will approach the width of the underlying uncertainty.

>>>>

I will also make a few  simplifying assumptions for this work. 
First, the systems are assumed to be 
time invariant. This 
assumption is based on the idea that if you paused the system for $\Delta t$
seconds, when unfrozen the system would continue as if nothing happened. 
Few biological systems are predictable enough for them to be summarized
by a time varying function, least of all the brain. While heart beats are certainly
periodic and have an effect on the BOLD signal, the period varies too much
for the system to be considered varying with time. 
Next, its assumes that input cannot directly
influence the output, which in the case of the BOLD signal is a good assumption.
Finally, because the only difference between the members of $X(t)$ and 
$\Theta$ is an update function, from now on $X$ will encapsulate
$\Theta$. The assumptions allow for a simplified version of the
state space equations:

\begin{eqnarray}
\dot{X}_k &=& f(X_{k-1}, u_k, \nu_x)
\label{eq:stateass}\\
Y_k &= &g(X_k, \nu_y)
\label{eq:measass}
\end{eqnarray}

@@

\begin{algorithm}
\caption{Resampling Algorithm}
\begin{algorithmic}
\STATE Calculate total weight, $W_t = \sum_{i=0}^{N_p} w^i$
\FORALL{$0 < i < N_p$}
    \STATE Draw $V$ from uniform range $[0, W]$
    \STATE $C = W_t$
    \FORALL{$0 < j < N_p$ and $C < V$}
        \STATE $C = C - w^j$
    \ENDFOR
    \STATE Add $[x^j, \frac{1}{N_p}]$ to the new distribution
\ENDFOR
\STATE 
\end{algorithmic}
\label{alg:Resampling}
\end{algorithm}
For infinite particles this new distribution will match the old.
Unfortunately, this isn't the truth in practice: since the support is
still limited to the original particles, the number of \emph{unique} particles can only go down.
This effect, dubbed particle impoverishment can result in excessive quantization
errors in the final distribution. However, there is a solution. Instead of sampling from the
discrete distribution, a smoothing kernel is applied, and particles are drawn from
that distribution. Because it is continuous, particle impoverishment
cannot occur. The easiest way to sample from the continuous distribution is to break the 
re-sampling down into two steps. After calculating an estimate of the scale of the original
distribution, algorithm \autoref{alg:Resampling} is performed. Next, a distribution is generated
based on the variance of the original distributions.
Finally, for each particle in the discretely re-sampled distribution, a sample is drawn from 
the smoothing 
distribution and added to the particle.  The regularization process is defined as:

@@

\STATE At $t + \Delta t$, $t \in T$, $P(x(t+\Delta t)) \approx 
\sum_{i=1}^{N_p} w_i(t)\delta\left(x - (x_i(t) + \int_t^{t+\Delta t} f(x(\tau), u(\tau)) d\tau) \right)$
 \end{algorithmic}
 \label{alg:RegularizedParticleFilter}
 \end{algorithm}

@@

With regularized resampling, it is possible to prevent both
particle degeneracy as well as particle impoverishment. 
Note that resampling carries certain risks.  
If for some reason the solution is not covered by the 
new support, the algorithm may not be able to reach the true value. 
The ultimate effect of this regularized resampling is a convergence 
similar to simulated annealing or a genetic algorithm. 
Versions of $x$ that are fit (give good measurements) spawn more children 
nearby which allow for more accurate estimation near points of high likelihood. 
As the variance of the estimated $x$'s decrease, the radius in 
which children are spawned also decreases. Eventually the radius
will approach the width of the underlying uncertainty.
